---
title: "esm206-notes-lab3"
author: "Danny Feinberg"
date: "10/18/2021"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(janitor)
library(here)
library(gapminder)
library(palmerpenguins)
```

### Part 0: Setup

- Create a new R project (does not need to be version-controlled)
- Add two subfolders: data and src
- Download two data files from GauchoSpace: jornada_lizards.csv and lizard_abb.csv, and copy them into your project data folder
- Open a new R Markdown document in your project, delete all below the setup chunk
- Attach the necessary packages in the setup chunk (tidyverse, janitor, here, gapminder). Note, you probably need to install the `gapminder` package by running `install.packages("gapminder")` in the Console

### Part 1: Tidying data continued

In this section, we'll learn how to combine two data frames with `full_join()`, and to reshape with `pivot_longer()` and `pivot_wider()` to get data into tidy format.

A. Join lizard common names with lizard data

The following data are from Jornada Basin LTER: 

- Title: Lizard pitfall trap data (LTER-II, LTER-III)
- URL: https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-jrn.210007001.36
- Citation: Lightfoot, D. 2017. Lizard pitfall trap data (LTER-II, LTER-III) ver 36. Environmental Data Initiative. https://doi.org/10.6073/pasta/ff37b4cace16a9943575f3fd7067064e (Accessed 2020-07-05).
- Lizard code list from: https://jornada.nmsu.edu/content/lizard-code-list

Read in both lizards datasets: 

- jornada_lizards.csv (pitfall trap data), which contains:
     - common name
     - site (site name)
     - sex (M = male, F = female, J = juvenile)
     - tail (B = broken, W = whole)
- lizard_abb.csv (lizard abbreviations & common names)


```{r}
# Read in lizards data
# Note: your path will differ depending on uyour project setup
lizard_names <- read_csv(here("data","lizard_abb.csv"))

lizard_traps <- read_csv(here("data","jornada_lizards.csv"))
```

Join together by the `common_name` variable with `full_join()`. See more information on `full_join()` - there are other options for more discriminating / excluding joins, but if you're not sure how they work **FULL JOIN** and deal with filtering later! 

```{r}
lizard_join <- lizard_traps %>% 
  full_join(lizard_names)
```

Notice that it will first look for matching names, and join by those. But we don't *have* to have matching names. For example, let's make a version of lizard_traps that has common_name renamed as "lizard_name":

```{r}
lizard_traps_rn <- lizard_traps %>% 
  rename(lizard_name = common_name)
```

We can then specify which column to join by: 

```{r}
lizard_join_2 <- lizard_traps_rn %>% 
  full_join(lizard_names, by = c("lizard_name" = "common_name"))
```

Refresh our memory on finding counts: 

```{r}
tail_counts <- lizard_join %>% 
  filter(common_name %in% c("Western Whiptail","Eastern Fence")) %>% 
  count(common_name, tail)

tail_counts
```

# Lab 3 Notes

## Exercise 1
```{r}
# find counts of lizards by sex and tail condition 
lizards_counts_1 <-  lizard_join %>% 
  group_by(sex, tail) %>% 
  summarize(sample_size = n())
  
lizards_counts_1 #show me the dataframe when knitting; don't just store it
```

### Using this same strategy find the number of lizards observed by SITE and COMMON NAME within this data fram (lizard_join)
```{r}
lizards_colunts_2 <- lizard_join %>% 
  group_by(site, common_name) %>% 
  summarize(count = n())

lizards_colunts_2
```

## Native pipe operator to R: `|>`
- most R functions should play nice with this standard pipe operator (i.e., you won't need to load in the tidyverse for it to work)

## Palmer Penguins example  
```{r}
penguins_n <- penguins %>% 
  group_by(species) %>% 
  mutate(sample_size = n()) %>% # tacks on a new column to each row indicating overall # sample size
  mutate(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE))
view(penguins)
```

## if you just want to find counts of lizard_join by common_name and tail
```{r}
lizard_count_3 <- lizard_join %>% 
  count(common_name, tail)

lizard_count_3
```


## recreate table above with group_by and summarize
```{r}
lizard_count_3_again <- lizard_join %>% 
  group_by(common_name, tail) %>% 
  summarize(n = n())

lizard_count_3_again
```
```{r}
mtcars

mtcars_summary <- mtcars %>% 
  group_by(cyl) %>% 
  summarize(mean_mpg = mean(mpg),
            sd_mpg = sd(mpg),
            sample_size = n())

mtcars_summary
```

  ## Putting together things that we've learned
  ### Starting from lizard_join, I want you to: 
    - create a subset only containing site, common_name, and tail status
    - Only including observations for "Western Whiptail", "Texas Horned", and "Little Striped"
    - for each remaining species, create a summary table that contains the counts of observations BY SPECIES (can do it however you want)
    
    #### My Attempt:
```{r}
my_attempt_lizard_join <- lizard_join %>% 
  select(site, common_name, tail) %>% 
  filter(common_name %in% c("Western Whiptail", "Texas Horned", "Little Striped")) %>% 
  count(common_name)

# Use uniuqe() to check all existing levels of a variable
#unique(my_attempt_lizard_join$common_name)
```
  
OR 

```{r}
my_attempt_lizard_join <- lizard_join %>% 
  select(site, common_name, tail) %>% 
  filter(common_name %in% c("Western Whiptail", "Texas Horned", "Little Striped")) %>% 
  group_by(common_name) %>% 
  summarize(total_count = n())

my_attempt_lizard_join
```
 ---
## How to reshape data to longer or wider formats
Sometimes we need to reshape data to get it into Tidy (TM) format. 

Tidy data: 
1. Each variable is in its own column 
2. Each observation is in its own row
3. Each value occupies a single cell 

(we are using the gapminder dataset for this exercise)

First make a subset with only observations from the US, Canada, and Mexico
```{r}
n_am <- gapminder %>%
  filter(country %in% c("United States", "Canada", "Mexico")) %>% 
  select(country, year, pop)
```

### New task: create a ggplot line graph, separated by country (from `n_am` subset) of population over time
```{r}
ggplot(data = n_am, aes(x = year, y = pop)) +
  geom_line(aes(color = country))
```


# to get information on a package: put a `?` in front of the package name
  - example: `?gapminder`
  

### Make a wide version of the gapminder data, where the year vairiable gets spread out across multiple columns


```{r}
n_am_wider <- n_am %>% 
  pivot_wider(names_from = year,
              values_from = pop) 

view(n_am_wider)

#pivot_wider is a function from tidyR package (part of tidyverse) 
```

#### To get it back to tidy format we want to collapse the year columns, then give the column the name "year", then put the values back in the correct columns

Now back to tidy format using `pivot_longer()`.

```{r}
n_am_longer <- n_am_wider %>% 
  pivot_longer(
    cols = '1952':'2007',
    names_to = 'year',
    values_to = 'pop'
  ) %>% 
 

#but `year` might show up as a character, rather than number as the variable type since it was a column name before.

mutate(year = as.numeric(year)) #converts existing year column (character) to numeric

class(n_am_longer$year) #checking class of new, converted year variable
```

## Make our first function
<mark>Shortcut: Cmd + Option + X</mark>

distance = rate * time 

```{r}
calc_distance <- function(rate, time) {
  rate * time
} 

calc_distance(rate = 1000, time = 42)

```

### new function 
```{r}
model_bananas <- function(A, B, C, bananas) {
  A + B - C / bananas
}

model_bananas(A = 1000, B = 2, C = 50, bananas = 0.0001)
```
