---
title: "esm206-notes-lab1"
author: "Danny Feinberg"
date: "10/18/2021"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

####------- Attach packages -------####
library(tidyverse)
library(janitor)
library(here)
```

## Objects Exercize at the beginning of class

```{r}
###############################################################
######    Objects exercise at the beginning of class    #######
###############################################################

# original prices
taco <- 6
price <- 3

# calculate meal price
price_meal <- taco * price

# change number of tacos to n=8
taco <- 8

# recalculate price_meal
price_meal <- taco * price

### making a vector (everything to the right of the arrow will be stored as "taco_week")    ###
### c() is a function that exists in base R; within parenthesis we pass the function values ###
### (in this case, the number of daily tacos for the last 7 days)                           ###

# vector describing number of tacos per day
taco_week <- c(2, 5, 10, 1, 4, 8, 2)

# functions we can compute based on vector above
mean(taco_week)

max(taco_week)

# store mean daily tacos as an object "mean_daily_tacos"
mean_daily_tacos <- mean(taco_week)
```

## Important Note!

**Do not put "install.packages()" code in an Rmarkdown doc because it will run every time you press the 'knit' button**

## Keyboard shortcuts

### To insert code chunk
- <mark>On a mac: "⌘" + "⌥" + "i"</mark> 
- On a PC: "ctrl" + "alt" + "i" 
- On either: click the green '+C' button above and choose 'R'

### To insert "<-" 
- <mark>On a mac: "⌥" + "-"</mark> 

### To insert pipe operator
- <mark>On a mac: ⌃⇧m ("ctrl" + "shift" + "m")</mark> 

## Use read_csv() - **DO NOT use read.csv**

```{r}
verts <- read_csv(here("data", "mack_creek_verts.csv"))
```

## Exploring the data

In the console, I can use the head() function to look at the first 6 lines - tail() does the last 6 lines - names() shows us the names

## Clean up the column names with janitor::clean_names()

-   <mark>control shift M</mark> is the shortcut for the **pipe operator "%\>%"**

    -   the pipe operator allows us to write code the way that our brains think about it (logically ordered rather than nested processes; "start with this *and then* do this thing)

-   clean_names() converts column names to lower snake case

```{r}
verts_clean <- verts %>% clean_names()
## Line above could have been written as follows: 
### clean_names(vert)
```

## Basic Data Wrangling using the dplyr package (exists within the tidyverse(a collection of packages))

-   <mark>use dplyr::select() to select/exclude columns; use dplyr::filter() to select/exclude rows</mark>

```{r}
verts_subset <- verts_clean %>% 
  select(species, length1, weight)
```

## Let's make our first gg plot

```{r}
ggplot(data = verts_subset, aes(x = length1, y = weight)) +
  geom_point()

# if you are referring to a column in ggplot, use aes (to map aes onto variable)
# the "+" sign adds a layer in ggplot
# geom_* give us many ways to plot
```

### Let's refine our plot with colors (add an aesthetic (aes(color = variable that should be identified with colors)))

```{r}

ggplot(data = verts_subset, aes(x = length1, y = weight)) +
  geom_point(aes(color = species))

```

## Let's make a subset of this data that only used information from the cutthroat species (ONCL in dataframe) of fish (filtering by ONCL species)

### Use dplyr::filter() to create subsets keeping rows based on whether they match our conditions or not.

```{r}
trout <- verts_clean %>% 
  filter(species == "ONCL")   #double equals sign means "must match"
```

### Piping multiple steps together

```{r}
verts_piped <- verts_clean %>% 
  clean_names() %>% 
  select(species, length1, weight) %>% 
  filter(species == "DITE")
  
```

```{r}
ggplot(data = verts_piped, aes(x = length1, y = weight)) +
  geom_point(aes(color = weight))
```

